---
title: "Population aging and COVID-19 death rate relationship analysis"
author: "Fanli Zhou"
date: "2020/5/6"
output: 
  github_document:
    toc: true
    pandoc_args: --webtex
bibliography: report_refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(mice)
library(resample)
library(repr)
library(cowplot)
library(knitr)
library(tidyverse)
```

```{r define functions and load data, echo=FALSE, message=FALSE}

get_min <- function(col) {
  round(min(data[[col]], na.rm = TRUE), 2)
}

get_max <- function(col) {
  round(max(data[[col]], na.rm = TRUE), 2)
}

data <- read_csv("../data/clean_data/complete_data.csv")

df_sub <- data %>%
    select(country, rate, age_65up, med_bed, con_500_ind, 
           death_100_ind, confirmed_rate) %>%
    drop_na()
```

# Summary



# Introduction

The world is fighting with the COVID-19. The COVID-19 is a new disease and we are updating our knowledge about it everyday. One well document factor that leads to death from COVID-19 is age. As reported by the US [CDC](https://www.cdc.gov/coronavirus/2019-ncov/need-extra-precautions/older-adults.html), adults 65 years old and older have high risk "for developing more serious complications from COVID-19 illness" [@cdc]. In other word, the COVID-19 death rate for the senior is higher than that for the younger people. This conclusion is based on the COVID-19 deaths statistics in the US. 

To further support this conclusion, I decided to study the relationship bewteen age and the COVID-19 death rate with data science techniques. Considering that COVID-19 patient information would not be public to protect privacy, I would instead study the relationship between population aging and the COVID-19 death rate. Here is my research question:

**Does the portion of the senior (age 65 and up) in the population (population aging) associate with the COVID-19 death rate?**

- Null hypothesis ($H_0$):

    + The portion of the senior (age 65 and up) in the population (population aging) is not associated with the COVID-19 death rate.

- Alternative hypothesis ($H_A$):

    + The portion of the senior (age 65 and up) in the population (population aging) is associated with the COVID-19 death rate.
    
This study would help address the relationship between age and the COVID-19 death rate. And hopefully, it would bring more attentions to protect the senior people from COVID-19 and protect ourselves.

# Methods

## Data

The two COVID-19 datasets, [the global confirmed dataset](https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv) and [the global death dataset](https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv), are from the [data repository](https://github.com/CSSEGISandData/COVID-19) for the 2019 Novel Coronavirus Visual Dashboard operated by the Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE). The dataset is updated everyday starting from January 22nd, 2020. Both datasets contain columns of location information (`Province/State`, `Country/Region`, `Lat`, and `Long`) or daily accumulated number of confirmed or deaths in the corresponding location, starting from `1/22/20` to present. As stated in the dataset repository, "Australia, Canada and China are reported at the province/state level. Dependencies of the Netherlands, the UK, France and Denmark are listed under the province/state level. The US and other countries are at the country level." The datasets combine data from multiple sources and may contain mistakes.

Here is information of variables used in this study:

|varaiable Name|Min|Max|Description and Source|
|-|-|-|-|
|`rate`|`r get_min("rate")`|`r get_max("rate")`|For each country, the death rate is calculated by $\frac{N_{\text{COVID-19 death}}}{N_{\text{COVID-19 confirmed}}}$ with data from [global confirmed](https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv) and [global death](https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv) datasets|
|`age_65up`|`r get_min("age_65up")`|`r get_max("age_65up")`|[Population ages 65 and above (% of total population)](https://data.worldbank.org/indicator/SP.POP.65UP.TO.ZS) from the [World Bank Open Data](https://data.worldbank.org/)|
|`air_pollution`|`r get_min("air_pollution")`|`r get_max("air_pollution")`|[PM2.5 air pollution, mean annual exposure (micrograms per cubic meter)](https://data.worldbank.org/indicator/EN.ATM.PM25.MC.M3) from the World Bank Open Data|
|`med_bed`|`r get_min("med_bed")`|`r get_max("med_bed")`|[Hospital beds (per 1,000 people)](https://data.worldbank.org/indicator/SH.MED.BEDS.ZS)|
|`physicians`|`r get_min("physicians")`|`r get_max("physicians")`|[Physicians (per 1,000 people)](https://data.worldbank.org/indicator/SH.MED.PHYS.ZS) from the World Bank Open Data|
|`nurses_midwives`|`r get_min("nurses_midwives")`|`r get_max("nurses_midwives")`|[Nurses and midwives (per 1,000 people)](https://data.worldbank.org/indicator/SH.MED.NUMW.P3) from the World Bank Open Data|
|`pop_density`|`r get_min("pop_density")`|`r get_max("pop_density")`|[Population density (people per sq. km of land area)](https://data.worldbank.org/indicator/EN.POP.DNST) from the World Bank Open Data|
|`smoking`|`r get_min("smoking")`|`r get_max("smoking")`|[Smoking prevalence, total (ages 15+)](https://data.worldbank.org/indicator/SH.PRV.SMOK) from the World Bank Open Data|
|`con_500_ind`|`r get_min("con_500_ind")`|`r get_max("con_500_ind")`|The ratio of days with fewer than 500 confirmed cases (accumulated) is calculated by $\frac{N_{\text{days with 1-500 confirmed cases since 1/21/2020}}}{N_{\text{days with more than one confirmed cases since 1/21/2020}}}$ with data from the [global confirmed](https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv) dataset|
|`death_100_ind`|`r get_min("death_100_ind")`|`r get_max("death_100_ind")`|The ratio of days with fewer than 100 deaths (accumulated) is calculated by $\frac{N_{\text{days with 1-100 deaths since 1/21/2020}}}{N_{\text{days with more than one deaths since 1/21/2020}}}$ with data from the [global death](https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv) dataset|
|`confirmed_rate`|`r get_min("confirmed_rate")`|`r get_max("confirmed_rate")`|The confirmed rate is calculated by $\frac{N_{\text{COVID-19 confirmed}}}{N_{\text{population}}}$ with data from the [global confirmed](https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv) dataset and the [population, total](https://data.worldbank.org/indicator/SP.POP.TOTL) dataset from the World Bank Open Data|
|`gov_resp`|`r get_min("gov_resp")`|`r get_max("gov_resp")`|This value is the highest stringency index for each country from the [Oxford Covid-19 Government Response Tracker data](https://raw.githubusercontent.com/OxCGRT/covid-policy-tracker/master/data/OxCGRT_latest.csv) from the [Oxford Covid-19 Government Response Tracker repository](https://github.com/OxCGRT/covid-policy-tracker/)|
|`ind_80`|`r get_min("ind_80")`|`r get_max("ind_80")`|This index is calculated by $\frac{N_{\text{days with 0-80 stringency index since 1/21/2020}}}{N_{\text{days since 1/21/2020}}}$ with data from the [Oxford Covid-19 Government Response Tracker data](https://raw.githubusercontent.com/OxCGRT/covid-policy-tracker/master/data/OxCGRT_latest.csv)|

## Analysis Methods

The multiple linear regression, GLM (General Linear Model) and bootstrapping algorithms or methods were used to study the relationship between the COVID-19 death rate and the portion of the senior. The Python programming language [@Python], the R programming language [@R] and the following Python/R packages were used: Pandas [@mckinney-proc-scipy-2010], requests, zipfile, io, json, knitr [@knitr], tidyverse [@tidyverse], MASS, mice, resample, testthat, rjson, repr, cowplot [@cowplot]. The code used to perform the analysis and create this report can be found here: https://github.com/flizhou/age_effects_on_COVID-19.

The following three functions were used to display model results and analysis.

``` {r}

#' Report linear regression results
#'
#' @param model the fitted model to analyze
#'
#' @return results dataframe
analyze_lm <- function(model){
    model %>%
        broom::tidy() %>%
        mutate(p.value.adjust = p.adjust(p.value, method = "BH"),
               significant = p.value.adjust < 0.05)
}

#' Plots linear regression diagnostics results
#'
#' @param model the fitted model to analyze
#' @param formula the model formula
#'
#' @return plots
plot_model_diag <- function(model, formula = "") {
    options(repr.plot.width = 20, repr.plot.height = 20)
    par(mfrow = c(2, 2), mar = c(5, 5, 4, 2) + 0.1, 
        cex.lab = 1.2, cex.axis = 1, cex.main = 1.2, cex.sub = 1.2)

    if (formula != "") {
        print(paste("Formula =", formula))
    }
    
    plot(model, cex.caption = 1.2, cex.id = 1)
}

t <- theme(plot.title = element_text(size = 16),
           axis.title = element_text(size = 14),
           text = element_text(size = 12),
           legend.title = element_text(size = 14),
           legend.text = element_text(size = 14),
           strip.text.x = element_text(size = 14))

#' Plots bootstrapping results
#'
#' @param boot_results the bootstrapping results
#' @param var the variable to plot
#' @param model_results the model results
#'
#' @return ggplot plot
plot_results <-  function (boot_results, var, model_results){
    
    boot_fits <- boot_results %>% 
        filter(term == var)
    
    obs <- model_results %>% 
        filter(term == var)
    
    t_star <- (boot_fits$estimate - obs$estimate) / boot_fits$std.error
    pval_boot <- round((1 + sum(abs(t_star) > abs(obs$statistic))) / (N + 1), 4)

    df <- tibble(estimate = boot_fits$estimate,
           t_star = t_star) %>%
        gather("term", "value") 
    
    df %>%
        ggplot() +
            geom_histogram(aes(value), bins = 30) +
            facet_grid(cols = vars(term), scales = "free") + 
            geom_vline(data = df %>% filter(term == "t_star"), 
                       aes(xintercept = obs$statistic, color = "Observed t")) +
            geom_vline(data = df %>% filter(term == "estimate"), 
                       aes(xintercept = obs$estimate, color = "Observed estimate")) +
            theme_bw() +
            labs(title = paste("Bootstrap distributions of the coefficient of ", var, 
                               "\np-value = ", pval_boot, sep = ""),
                 color = "Line") + t
    
}
```

# Results and Discussion

## 1. multiple linear regression

I started my analysis with a multiple linear regression model to study the relationship between the COVID-19 death rate ($Y$, rate) and the portion of the senior ($X$, age_65up). A multiple linear regression model may not give me the most accurate model but is highly interpretable. To answer my research questions, I would sacrifice some accuracy for interpretability.

I first tried the full model, which includes all confounding variables, and picked out potential useful confounding variables based on statistic significance (p-value < 0.05). Then I tested reduced models to decide which variables to include in the final model. Considering that I was doing multiple testing, I controlled for multiple comparisons using Benjamini & Hochberg (BH) False Discovery Rate (FDR) adjustment (setting FDR to 5%). Based on these analysis, my final model only includes `med_bed`, `death_100_ind`, and `confirmed_rate` as confounding variables.

```{r data}
# load the complete dataset that includes all confounding variables
data <- read_csv("../data/clean_data/complete_data.csv")

# remove confounding variables that won't be used in the final model
df_sub <- data %>%
    select(country, rate, age_65up, med_bed, 
           death_100_ind, confirmed_rate) %>%
    drop_na()
head(df_sub)
```

```{r model 1}
# the final multiple linear regression model
model_1 <- lm(rate ~ age_65up + med_bed + death_100_ind + confirmed_rate, df_sub[-1])

analyze_lm(model_1)
```

To assess this model, I checked whether the "LINE" conditions for multiple linear regression hold. The data were collected in each country independently, so I would assume that the independence assumption is reasonable. Then, let's look at the rest assumptions:

```{r model 1 analysis}
# plot model diagnostics
plot_model_diag(model_1, "rate ~ age_65up + med_bed + death_100_ind + confirmed_rate")

# Compute Shapiro-Wilk test of normality
rstatix::shapiro_test(residuals(model_1))

# Compute Bartlett test of homogeneity of variances
df_sub[-1] %>%
    bartlett.test(rate ~ age_65up + med_bed + death_100_ind + confirmed_rate)
```

First, check the Residuals vs Fitted plot for the linearity assumption. Residuals should have zero expectations at each fitted value. There are some obvious patterns in the plot and the averages of residuals are away from zero. So the linearity assumption does not hold and the LSE (Least Squares Estimator) might be biased and even inconsistent.

Second, check the Normal Q-Q plot for the normality assumption. Deviations of the lower and upper tails from the straight line indicates that the distribution of residuals does not conform to a normal distribution (right skewed in this case). Besides, the Shapiro-Wilk test of normality is significant. So the normality assumption is violated.

Third, check the Scale-Location plot for the equal variance assumption. There is a overall positive trend in the plot so the variance is not constant. Besides, the Bartlett test of homogeneity of variances is significant. So the equal variance assumption does not hold and the LSE may not be the BLUE (Best Linear Unbiased Estimator).

Finally, check the Residuals vs. Leverage plot for influential points. Three point are labeled for high Cook's distance values.

The model violates most of the "LINE" conditions. I next considered improving the model with data transformation and GLM (General Linear Model).

## 2. Data transformation and GLM

I tried to improve the model by trial and error. Here is the final model I got. The final model was the best model identified, but not necessary the best model.

```{r model 2}
model_2 <- df_sub %>%
    mutate(rate = (rate*(133 -1) + 2)/133) %>%
    glm(rate ~ log(age_65up) + log(med_bed) + 
        log(death_100_ind + 1) + log(confirmed_rate), 
        family = Gamma(link = "identity"), .)

analyze_lm(model_2)
```

Again I checked whether the "LINE" conditions hold to assess the model. The independence assumption is reasonable as before. For the rest assumptions:

```{r model 2 analysis}
# plot model diagnostics
plot_model_diag(model_2)

# Compute Shapiro-Wilk test of normality
rstatix::shapiro_test(residuals(model_2))

# Compute Bartlett test of homogeneity of variances
df_sub[-1] %>%
    mutate(rate = (rate*(133 -1) + 2)/133) %>%
    bartlett.test(rate ~ log(age_65up) + log(med_bed) + log(death_100_ind+1) + log(confirmed_rate))
```

First, the averages of residuals are slightly smaller than zero in the Residuals vs Fitted plot but may not be a big problem. I would say that the linearity assumption could hold here.

Second, the lower and upper tails are slightly away from the straight line in the Normal Q-Q plot. Besides, the Shapiro-Wilk test of normality is significant but the p-value is very close to 0.05. So there is some evidence that the normality assumption could be reasonable.

Third, the Scale-Location plot has some trends so the variance is not constant. And the Bartlett test of homogeneity of variances is significant. So the normality assumption could not hold.
Finally, three point are labeled for high Cook's distance values in the Residuals vs. Leverage plot.

The final model still violates some of the "LINE" conditions. So it may be not be reasonable to use the theoretical inferential conclusions. Instead, I used bootstrapping as an alternative approach to statistical inference.

## 3. Bootstrapping

I took bootstrap samples from 133 countries in `df_sub` with replacement. This was case resampling, so the information from each country remained together. Here I had to drop the `log(confirmed_rate)` term in the formula to make the `glm` results converge.

```{r model 3}
set.seed(12345)
N <- 1000

model_3 <- df_sub %>%
    mutate(rate = (rate*(133 -1) + 2)/133) %>%
    glm(rate ~ log(age_65up) + log(med_bed) + 
        log(death_100_ind + 1), 
        family = Gamma(link = "identity"), .)

model_3_results <- analyze_lm(model_3)

boot_results <- df_sub %>%
    mutate(rate = (rate*(133 -1) + 2)/133) %>%
    rsample::bootstraps(N) %>% 
    mutate(lm = map(splits, ~ glm(rate ~ log(age_65up) + log(med_bed) + 
                                  log(death_100_ind + 1), 
                                  family = Gamma(link = "identity"), .)),
           tidy = map(lm, broom::tidy)) %>% 
    select(-splits, -lm) %>% 
    unnest(tidy)

plot_results(boot_results, "log(age_65up)", model_3_results)
plot_results(boot_results, "log(med_bed)", model_3_results)
plot_results(boot_results, "log(death_100_ind + 1)", model_3_results)
```


Based on the bootstrapping results, the coefficients of `log(age_65up)`, `log(med_bed)`, `log(death_100_ind + 1)` are all significant (p-value < 0.05).

## Conclusions

In this analysis, I attempted to use a linear model to explain the relationship between the portion of the senior (`age_65up`%) in the population and the COVID-19 death rate. I tried several linear models and finally decided to train a generalized linear model with a Gamma-distribution dependent variable. My final model is:

$\frac{132\text{rate}+2}{133}$ $=\0.103+\0.0133\text{log(age\_65up)}-\0.0149\text{log(med\_bed)}-\0.103\text{log(death\_100\_ind+1)}$

where the response (rate) follows a Gamma distribution.
I understand that this is a simplified model and I may not include all possible confounding variables in the model. And the "LINE" conditions for multiple linear models may not hold for this model. So I used bootstrapping for statistical inference. Given the bootstrapping results, the coefficient of `log(age_65up)` is significant (p-value < 0.05). I have the evidence to reject the null hypothesis and accept the alternative hypothesis. So the portion of the senior (age 65 and up) in the population (population aging) is associated with the COVID-19 death rate.

# Limitations
 
# References